---
title: "TB prevalence in wild boar"
author: "Zara Baker"
date: "2024-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TB prevalence in wild boar

This dataset contains information on the disease phenotype of wild boar and investigates the relationship between this and potentially associated variables of sex, length and age.

### Table to explain our variables:

| Variable | Type of Variable | Description |
|:-----:|:---------:|:---------:|
| TB | Binary | Categorising individuals as infected (1) or uninfected (2) |
| Age | Categorical | defining individuals into 1 of 4 age categories - juvenile, subadult, young adult, old adult (accurate age cannot be clearly defined in years) |
|Length | Continuous | Nose-rump length, measured in cm |
| Sex | Categorical | Defining individuals as male (1) or female (2) |

### The Aim:
Identify the risk factors for TB in this population. E.g. is infection equally distributed among the population? Can we make predictions about likelihood?

### Step 1: Reading in the data

This command will read in the data (provided you are within an R project), assigns NAs, and converts strings to factors:

```{r Step 1: read in}
dframe1<-read.csv("data/Boar.csv",na="NA",stringsAsFactors =T)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
